Pareto-Optimal Multistate Protein Design using Tree Graphical Models

### 2012-12-27 docs/tex/draft.tex
Finished Simulation 1, sim 2 debugging and results, sim3 skeleton. Added changes to document

\subsubsection{Training an HMM}

\subsubsection{Converting a HMM to CMRF}
It is possible to reparameterize an HMM so that it now becomes a CMRF. The partition function $Z$ of the newly constructed CMRF equals 1 corresponding to a probability distribution.

\subsection{Combining CMRFs to get a TMRF}

@@

\subsection{Simulation Experiments}
We demonstrate how our algorithm performs on simulated data. The data was prepared by considering a protein of length 12. The sequence space was limited to an alphabet consisting of two arbitrary characters $\alpha,\beta$. The observation belong to a discrete space Helix, Beta and Loop which represented by H,B and L. The weights of the CMRF have weights $\phi,\psi$ as defined in Tables~\ref{tab:psi1} and~\ref{tab:psi2}. We define two states State A and State B that corresponding to competing objectives for multistate design. State A is a Helix-Loop-Helix strand whereas State B is Sheet-Loop-Sheet strand. Figure~\ref{fig:simstates} shows a cartoon diagrams for our design problem. State A and State B have \texttt{HHHHLLLLHHHH} and \texttt{BBBBLLLLBBBB} secondary structure features respectively. These are the observations for the CMRF.
\\
\\
Using the weight matrices we defined in Tables~\ref{tab:psi1} and~\ref{tab:psi2} we run CMRF-Decode to get $X_A$ which satisfies $X_A = \argmin_X E_A(X)$. We obtain $X_B$ similarly. Not surprisingly $X_A$ is $\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha$ 
and $X_B$ is $\beta\beta\beta\beta\beta\beta\beta\beta\beta\beta\beta\beta$. This is because the weights in the table are biased towards $\alpha$ emitting \texttt{H} and $\beta$ emitting \texttt{B}. Thus according to our model $X_A$ is the sequence that is most stable under state A, whereas $X_B$ is the sequence is most stable under state B.
\\
\\
We then run the \texttt{Pareto-Frontier} algorithm using the same weights and observations for state A and state B to enumerate the entire pareto frontier. This pareto frontier contains all the best sequences that any convex combination of $E_A(X)$ and $E_B(X)$ would give rise to. We number the points to show the path taken by the pareto frontier algorithm. We also calculate the energies of all the $2^{12}$ sequences possible under this restricted sequence model and plot the points. The \texttt{Pareto-Frontier} algorithm was able to find the pareto frontier in \textcolor{red}{N steps} steps. See Figure~\ref{fig:pareto-sim} for the trace taken by the algorithm.  We also compare the running times as opposed to brute force and MCMC and compare the running times.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{pics/convexfront2.pdf}
    \caption{ }
    \label{fig:pareto-sim}
\end{figure}
\pagebreak

@@

@@

\paragraph{Multiple sequence Same Energy} :  Currently the cvhull algorithm is being thrown out of gear by sequences that have the same energy. An example of this situation is shown in Figure~\ref{fig:sim2_debug1}
\begin{figure}[h!]
    \centering
    \includegraphics[width=1.0\textwidth]{pics/sim2_debug1.pdf}
    \caption{This figure illustrates the issue when there are multiple sequences with the same energy. Here sequences 3 and 4 have the same energy. This causes the convex hull algorithm to fail in its current implementation.   }
    \label{fig:sim2_debug1}
\end{figure}
\pagebreak

>>>>

\subsubsection{Training an HMM}
Training an HMM requires the following 
\begin{itemize}
\item Protein sequence alignments
\item A mapping between amino acids of the sequences to discrete features
\end{itemize}

The emission and transition probabilities are calculated as

\begin{align*}
P(X_{i+1}=b|X_i=a) = \frac{\text{count}(X_{i+1}=b)+d}{\text{count}(X_i=a)+dS} \\
P(O_{i}=f|X_i=a) = \frac{\text{count}(O_{i}=f)+d}{\text{count}(X_i=a)+dF}
\end{align*}
where d is a smoothing count and $S$ and $F$ are the sizes of the alphabet of the sequence space and the feature space respectively.

\subsubsection{Converting a HMM to CMRF}
It is possible to reparameterize an HMM so that it now becomes a CMRF. The partition function $Z$ of the newly constructed CMRF equals 1 corresponding to a probability distribution.
\\
\\
This can be done as :
\begin{align*}
\phi(X) &= 1 \\
\phi(\mathbf{O_i})&=1 \\
\psi(X_i,X_{i+1}) &= P(X_i,X_j)\\
\psi(X_i,\mathbf{O_i}) &= P(\mathbf{O_i}|X_i)
\end{align*}

\subsection{Combining CMRFs to get a TMRF}
Instead of having different nodes for all the the different TMRFs. The energy functions can be so modified such that they correspond to a single CMRF. then decoding and sampling from this CMRF is the same as before. 
\\
\\
This can be done as 
\begin{align*}
\phi^{TMRF}(X) &= 1 \\
\phi^{TMRF}(\mathbf{O_i}) &= 1 \\
\psi^{TMRF}(X_i,X_{i+1}) &= \psi(X_i,X_j)\\
\psi^{TMRF}(X_i,\mathbf{O_i^{AB}}) &= \psi(X_i,\mathbf{O_i^A})^{\theta_A}\psi(X_i,\mathbf{O_i^B})^{\theta_B}
\end{align*}

\subsection{Augmenting Feature Space}
The feature space can be augmented to include any set of discrete features $F_1,F_2,\cdots,F_k$. Such a feature space can be added on to the existing feature space using a tree-like structure as shown in Figure~\ref{fig:CMRF_augmented}. 
\\
\\
The probability of a sequence assignment is defined as follows :
\begin{align*}
P(X_1,X_2,\dotsc,X_n) &= \frac{1}{Z}\prod_{i=1}^{n} \phi(X_i)\left(\prod_{j=1}^{k}\phi(\mathbf{O_{iF_j}})\psi(\mathbf{O_{iF_j}},X_i)\right) \prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})} 
\end{align*}

\begin{figure}[h]
% selecting unit length
\centering
% used for centering Figure
\setlength{\unitlength}{0.3cm}
\begin{picture}(40,20)(0,0)
\multiput(7,15)(8,0){1}{\circle{4}}
\multiput(3,7)(8,0){1}{\circle{4}}
\multiput(3,7)(8,0){1}{\circle{3.7}}
\multiput(11,7)(8,0){1}{\circle{4}}
\multiput(11,7)(8,0){1}{\circle{3.7}}
\thicklines
\multiput(9,15)(8,0){1}{\line(1,0){10}}
\multiput(20,15)(1,0){4}{\circle*{0.3}}
\multiput(7,13)(8,0){1}{\line(-1,-1){4}}
\multiput(7,13)(8,0){1}{\line(0,-1){4}}
\multiput(7,13)(8,0){1}{\line(1,-1){4}}
\multiput(6,7)(1,0){3}{\circle*{0.3}}
\put(24,15){\line(1,0){4}}
\put(30,13){\line(-1,-1){4}}
\put(30,13){\line(1,-1){4}}
\put(30,13){\line(0,-1){4}}
\multiput(29,7)(1,0){3}{\circle*{0.3}}
\thinlines
\put(30,15){\circle{4}}
\put(26,7){\circle{4}}
\put(26,7){\circle{3.7}}
\put(34,7){\circle{4}}
\put(34,7){\circle{3.7}}
\put(5.5,14.6){$\phi(X_1)$}
\put(28.5,14.6){$\phi(X_n)$}
\put(0.5,6.6){$\phi(O_{1F_1})$}
\put(8.5,6.6){$\phi(O_{1F_k})$}
\put(23.5,6.6){$\phi(O_{nF_1})$}
\put(31.5,6.6){$\phi(O_{nF_k})$}
\put(22.5,16.5){\footnotesize{$\psi(X_{n{-}1}{,}X_n)$}}
\put(-1,10.6){\footnotesize{$\psi(X_1{,}O_{1F_1})$}}
\put(10,10.6){\footnotesize{$\psi(X_1{,}O_{1F_k})$}}
\put(33.2,10.6){\footnotesize{$\psi(X_n{,}O_{nF_1})$}}
\put(21.2,10.6){\footnotesize{$\psi(X_n{,}O_{nF_k})$}}
\end{picture}
\label{fig:CMRF_augmented}
\caption{A CMRF with augmented feature space. Note the tree like shape of the feature space. Here $F_1,F_2,\dots,F_k$ correspond to different discrete features. }
\end{figure}

@@

\subsection{Simulation Experiments}
We run a variety of different simulation experiments to check the performance of our algorithm on

\subsubsection{Sim Experiment 1}
We demonstrate how our algorithm performs on simulated data. The data was prepared by considering a protein of length 12. The sequence space was limited to an alphabet consisting of two arbitrary characters $\alpha,\beta$. The observation belong to a discrete space Helix, Beta and Loop which represented by H,B and L. The weights of the CMRF have weights $\phi,\psi$ as defined in Tables~\ref{tab:psi1} and~\ref{tab:psi2}. We define two states State A and State B that corresponding to competing objectives for multistate design. State A is a Helix-Loop-Helix strand whereas State B is Sheet-Loop-Sheet strand. Figure~\ref{fig:simstates} shows a cartoon diagrams for our design problem. State A and State B have \texttt{HHHHLLLLHHHH} and \texttt{BBBBLLLLBBBB} secondary structure features respectively. These are the observations for the CMRF.
\\
\\
Using the weight matrices we defined in Tables~\ref{tab:psi1} and~\ref{tab:psi2} we run CMRF-Decode to get $X_A$ which satisfies $X_A = \argmin_X E_A(X)$. We obtain $X_B$ similarly. Not surprisingly $X_A$ is $\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha\alpha$ 
and $X_B$ is $\beta\beta\beta\beta\beta\beta\beta\beta\beta\beta\beta\beta$. This is because the weights in the table are biased towards $\alpha$ emitting \texttt{H} and $\beta$ emitting \texttt{B}. Thus according to our model $X_A$ is the sequence that is most stable under state A, whereas $X_B$ is the sequence is most stable under state B.
\\
\\
We then run the \texttt{Pareto-Frontier} algorithm using the same weights and observations for state A and state B to enumerate the entire pareto frontier. This pareto frontier contains all the best sequences that any convex combination of $E_A(X)$ and $E_B(X)$ would give rise to. We number the points to show the path taken by the pareto frontier algorithm. We also calculate the energies of all the $2^{12}$ sequences possible under this restricted sequence model and plot the points. The \texttt{Pareto-Frontier} algorithm was able to find the pareto frontier in \textcolor{red}{1 step} steps. See Figure~\ref{fig:pareto-sim} for the trace taken by the algorithm.  We also compare the running times as opposed to brute force and MCMC and compare the running times.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{pics/sim1.pdf}
    \caption{Simulation Experiment 1: The emission and transition weights are as defined in Tables~\ref{tab:psi1} and~\ref{tab:psi2}. The Pareto-HMM finds the optimal pareto frontier for a twelve length sequence using the alphabet \texttt\{a,b\} and the feature space \texttt\{H,L,B\} in two steps of the Convex hull algorithm. The magenta stars corresponds to the optimal sequences. The magenta line corresponds to the lower left convex hull of the sequence energy space. The blue stars correspond to non-optimal dominated sequences.  The Pareto-Frontier algoerithm provably finds the frontier quickly instead of having to enumerate,calculate and sort the energies of a  $2^{12}$ sequence space.}
    \label{fig:pareto-sim}
\end{figure}

\subsubsection{Simulation Experiment 2}
In this example we look at a slightly more complicated energy frontier. To generate this energy surface we resorted to generating random emission and transition probabilities till we found a sufficiently complicated energy frontier but still having alphabet \texttt{\{a,b\}} and feature space \texttt{\{H,L,B\}}. The probabilities for this simulation experiment are shown in tables~\ref{tab:psi3} and~\ref{tab:psi4}.

\begin{table}
\begin{center}
\begin{tabular}{cc|c|c|c|}
\cline{3-5}
& & \multicolumn{3}{c|}{Observations($\mathbf{O_i}$)} \\ \cline{3-5}
& & Helix(H) & Beta(B) & Loop(L) \\ \cline{1-5}
\multicolumn{1}{|c|}{\multirow{2}{*}{Sequence($X_i$)}} &
\multicolumn{1}{|c|}{$\alpha$} & 0.334 & 0.272 & 0.394      \\ \cline{2-5}
\multicolumn{1}{|c|}{}                        &
\multicolumn{1}{|c|}{$\beta$} & 0.477 & 0.093 & 0.430      \\ \cline{1-5}
\end{tabular}
\end{center}
\caption{$\psi(X,\mathbf{O})$ contains the weights for edge potentials of sequence and features together }
\label{tab:psi3}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{cc|c|c|}
\cline{3-4}
& & \multicolumn{2}{c|}{Sequence($X_{i+1}$)} \\ \cline{3-4}
& & $\alpha$ & $\beta$  \\ \cline{1-4}
\multicolumn{1}{|c|}{\multirow{2}{*}{Sequence($X_i$)}} &
\multicolumn{1}{|c|}{$\alpha$} & 0.483 & 0.517      \\ \cline{2-4}
\multicolumn{1}{|c|}{}                        &
\multicolumn{1}{|c|}{$\beta$} & 0.589 & 0.411      \\ \cline{1-4}
\end{tabular}
\end{center}
\caption{$\psi(X_i,X_{i+1})$ contains the weights for the edge potentials of adjacent amino acids}
\label{tab:psi4}
\end{table}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{pics/sim2_all.pdf}
    \caption{Simulation Experiment 2: Pareto Frontier algorithm is applied to a more complicated energy surface. The algorithm finds the pareto frontier denoted by the dashed magenta in 5 steps, which is significantly faster than brute force enumeration.  }
    \label{fig:sim2_all}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{pics/sim2_frontier.pdf}
    \caption{Simulation Experiment 2(Frontier): This shows the results from simulation 2. The sequences along the frontier are shown as well }
    \label{fig:sim2_frontier}
\end{figure}

\pagebreak
\clearpage

@@

\subsubsection{Continuous Feature Spaces}
Currently we are unable to provide support for continuous feature spaces.  Performing inference on discrete spaces by conditioning on continuous feature spaces can be more challenging.

@@

\paragraph{Multiple sequence Same Energy} :  Currently the cvhull algorithm is being thrown out of gear by sequences that have the same energy. An example of this situation is shown in Figure~\ref{fig:sim2_debug1}. The current way we are dealing with this is to define a taboo list of energies that are to be ignored if they are found. A better way to do this would be to find all sequences with the same energy during the decoding stage and still use the same energy taboo list.

\begin{figure}[h!]
    \centering
    \includegraphics[width=1.0\textwidth]{pics/sim2_debug1.pdf}
    \caption{This figure illustrates the issue when there are multiple sequences with the same energy. Here sequences 3 and 4 have the same energy. This causes the convex hull algorithm to fail in its current implementation.   }
    \label{fig:sim2_debug1}
\end{figure}
\pagebreak
